Detection of eco-evolutionary dynamics in metacommunities using Joint Species

### 2024-11-30 manuscript/manuscript_v13.Rmd
Final draft manuscript, not knit

Biodiversity at the metacommunity scale is typically influenced by a number of environmental, spatial, biotic, and stochastic factors. At the same time, these factors impact the evolution of individual species, as sites present different local selection pressures and connectivity can impact gene flow and genetic drift. Identifying the relative impacts of environmental, spatial, biotic, and other drivers on community composition across spatial and temporal scales has been greatly facilitated by joint species distribution models, but these models have yet to consider the impact of microevolution on community composition. We used Heirarchical Models of Species Communities (HMSC) to analyze simulated data of an populations and communities, including in a large evolving metacommunity model, to establish whether HMSC can sufficiently quantify the contribution of phenotypic evolution for metacommunity composition. The models successfully partitioned variance contributed by environmental, spatial, and evolving phenotypic drivers, and also estimated site- and year-specific covariance. We also applied the HMSC with trait evolution model to an existing dataset studying trait change and community dynamics in experimental aquatic plant system. The study of eco-evolutionary dynamics may require data that reflects numerous complex, interacting processes and it is necessary to have flexible, generalized statistical models to analyze this data. JSDM models such as HMSC present one promising path for analysis of eco-evolutionary dynamics in multi-species communities.

@@

We evaluated whether linear regression models can appropriately capture the dynamics that drive species abundances in relatively simple populations and communities. We fit data generated from simulation models with increasing complexity (Table \@ref(tab:tab1)). to Bayesian linear regression models. These simulation models were (A) a single-species model of discrete-time logistic population growth, (B) expanded to include a randomly fluctuating environmental property $E$ that impacts the species' population growth rate, (C) a two-species model of logistic growth and competition that was then (D) expanded to consider a randomly fluctuating environmental property, (E) and further expanded to consider evolution in a trait $x$ that is selected by the environmental property, and finally (F) the model of two-species growth, competition, environmental change, and evolution in a spatially structured (10-patch) habitat.

@@

To fit linear regression models to the simulated population size data, we began with models using population size (model *A*) and environment (including the second-order term, to better reflect species optimum values; model *B*) as fixed effects ($N_t \sim N_{t+1} + E_t + E_t^2$), and progressed to a mixed model using environment and trait evolution as fixed effects ($N_t \sim E_t + E_t^2 + |\Delta x|$; for simulation model *E*) and time (models *C*, *D*, *E*, and *F*) and spatial distance (model *F*) as random effects. For consistency, all linear regression models were implemented as hierarchical Bayesian models using HMSC (Hierarchical Modelling of Species Communities [@Ovaskainen2017HMSC] [@Tikhonov2020]). HMSC uses a latent variable approach as an alternative to directly modelling species interaction coefficients, estimating a reduced number of linear combinations of species abundances that best predict future population size values. For linear models fit to data generated in simulations *C*, *D*, *E*, and *F* above, the time random effect was implemented to capture species interactions (as part of the temporal species association matrix estimated by the model), and the spatial random effect was implemented in model *F* to capture spatial covariance in abundances as a function of distance between sites. Full methodological details are given in Appendix S1. All simulations were run in R (@R2024) and HMSC models were implemented using the `Hmsc` package (@Hmsc2022).

@@

All simulations used the same parameter values ($W_{max}$ = 2, *P* = 1, *w* = 2, $N_c$ = 100, *p* = 0.001), species interaction matrix ($\alpha_{ii} = 0.00125$; $\alpha_{ij} \sim U(0,0.0015)$), and initial conditions (initial richness $s_{0k} \sim Pois(0.75)$; initial population size $N_{0i} \sim Pois(10)$; initial degree of maladaptation $\beta_{i0} \sim Gamma(0.75,1)$, initial distance to the local patch's optimum phenotype $d_{i0}=\sqrt{B_{i0}(w+P)}$, initial trait value $x_{i0} = E_k - d_{i0}$; Appendix S4). Simulations were run across a range of dispersal values (identical for all species; *d* = 0, $10^{-9},10^{-8},10^{-7},10^{-6},10^{-5},10^{-4},10^{-3},10^{-2}$, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) and heritability values (identical for all species: $h^2$ = 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) for 200 time steps, producing records of $N_{ikt}$ and $x_{ikt}$.

@@

To determine the relative influence of environmental and spatial properties, intra- and interspecific population dynamics, and evolutionary dynamics for community structure, we applied hierarchical modelling of species communities (HMSC, [@Ovaskainen2017HMSC] [@Tikhonov2020]) to model the log-abundance at each site and year for each species. The log-abundance values were modeled as $ln(N_{ikt}) \sim N(L_{ikt},\sigma_2)$, where the linear predictors **L** are the sum of fixed and random effects **L^F^** + **L^R^** [@Tikhonov2020]. The fixed effects were $E_k$, $E_k^2$, and the absolute value of the change in each species' trait value from the previous time step to the next $|\Delta x_{ik,t-1 \to t}|$. Random effects included time (the day of the sample) and space. The spatial latent variables represent a spatial model, where species respond to some latent spatial predictor associated with the *xy* coordinates of the sites [@Tikhonov2020]. When there was no dispersal in the system ($d=0$), spatial random effects were not included in the model.

### Applying HMSC on empirical data

To test the applicability of HMSC on ecological and evolutionary effects in empirical data, we applied HMSC on the experimental data of Jewel & Bell (2023) [@Jewell2023]. The authors conducted a community selection experiments in which they grew four floating and short-lived aquatic plants (the angiosperms *Lemna minor* here designated as Lm, *Spirodela polyrhiza* designated as Sp and *Wolffia columbiana* designated as Wc, and the liverwort *Ricciocarpos natans* designated as Rn) under different environmental conditions for twelve weeks. The environmental conditions were defined by three different shade cover levels for the experimental pools, 3%, 12% and 50%, and different nutrient concentrations, with low (dissolved nitrogen (DN) = 200 $\mu g L^{-1}$, dissolved phosphorous (DP) = 10 $\mu g L^{-1}$, medium (DN = 800 $\mu g L^{-1}$, DP = 40 $\mu g L^{-1}$), and high (DN = 3200 $\mu g L^{-1}$, DP = 160 $\mu g L^{-1}$) concentrations of nutrients. They combined each shade and nutrient level, resulting in nine different environmental conditions. For the eleven weeks of the experiment, the frond size of ten individuals from each species, as well as population density were measured. To use the data with our HMSC analysis, we calculated the average trait change as $\Delta x = x_{ijt} - x_{ijt-1}$, where $x_{ij}$ described the mean trait of species $i$ in environment $j$ and week $t$.

To determine the relative importance of evolution for community composition, we fitted two HMSC models to the data: An 'evolution model' in which $|\Delta x|$ of each species and environmental conditions were used as explanatory variables, and a 'no-evolution model', in which only the environmental conditions were included. Species density was log-transformed species before applying it to the model, therefore we could use normal distribution in the model. We also included replicates (*n* = 2) and time as random effects. For each model we ran four MCMC chains to generate 1000 samples of the posterior distribution for all model parameters. To evaluate the model-estimated role of evolution for predicted population size, we used the posterior distributions of the estimated effect size of $|\Delta x|$ multiplied by the observed $\Delta x$ of each species, added to the model-estimated intercept.

@@

Researchers often use linear regression models to study non-linear population and community dynamics. We evaluated the performance of linear models in simple (1-2 species, 1-few sites) cases, to validate that effect size estimates and predicted densities accurately reflected mechanisms before moving onto more complex models in the next sections of our analysis. When only one species was present (models *A* and *B*), the regression coefficients (model *A*: $N_{t+1} \sim \beta_0 + \beta_1 N_t + \epsilon$; *B* $N_{t+1} \sim \beta_0 + \beta_1 N_t + \beta_2 E_t + \beta_3 E_t^2 + \epsilon$) reflected the direction and magnitude of the mechanistic drivers of population size (i.e. $\beta_1$ correctly captured the positive relationship between $N_t$ and $N_{t+1}$ under our simulation conditions, and $\beta_2$ and $\beta_3$ correctly captured the quadratic unimodal relationship between $E_t$ and $N_t+1$; Appendix S1.1, S1.2).

When a second species was added (models *C* and *D*), the regression coefficients performed well for predicting density values at all time points, but the direction and magnitude of the regression coefficients did not reflect the intra- and interspecific competition strengths included in the mechanistic model (Appendix S1.3). We instead used a latent variable modelling approach (implemented in HMSC) to consider these species interactions as part of the temporal random effects included in the model. This approached produced effect size estimates that both predicted the population dynamics well (Figure \@ref(fig:fig1) a) and captured the conditional effects of environment for species densities (Figure \@ref(fig:fig1) b). The variation partition also successfully captured the relative importance of the environmental predictor and the temporal random effects for model *D* (which in this simple case were driven by intra- and interspecific species interactions; Figure \@ref(fig:fig1) c; Appendix S1.3, S1.4).

Trait evolution was included as a predictor when it was introduced in model *E*. The resulting linear regression coefficients could predict species dynamics (Figure \@ref(fig:fig1) d) and reflected the impact and magnitude of environmental and trait evolution predictors (Figure \@ref(fig:fig1) e). The variation partition also showed that trait evolution in species 1 ($|\Delta x_1|$) was a strong driver of variation in abundance (blue portion, Figure \@ref(fig:fig1) f), primarily for species 1 but also for species 2 as well (Appendix S1.5; Appendix S2).

After adding spatial complexity to the 2-species simulation, we included a spatial random effect to capture species covariances as a function of distance between sites (model *F*). By properly considering species interactions as a temporal random effect and including spatial random effects, the HMSC model could estimate regression coefficients for the impacts of the environment $E_t$ and trait evolution $|\Delta x_1|$ and $|\Delta x_2|$ for both species that fit well to the simulated data points (Figure \@ref(fig:fig1 g); Appendix S1.6) and capture conditional effects of environment and trait evolution for both species (Figure \@ref(fig:fig1) h). Spatial random effects explained a large proportion of variance in abundance for both species, but trait evolution remained a strong driver for species 1.

```{r fig1, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("Simulation models for 2-species ecological and eco-evolutionary dynamics. Panels a, d, and g show log-population size (y-axis) over time (x-axis) for species $1$ and $2$ when they grow and compete for resources with random environmental fluctuations over time. Species either have fixed (panel a) or evolving trait values (panels d, g) and are present in either 1 (panels a, d; simulation models D and E) or multiple sites (panel g; model F). Graphs in panels b, e, and h show predicted log-population size ($ln(N)$) for species 1 and 2 across gradients of the environment ($E + E^2$) and the change in trait values of species 1 ($|\\Delta x_1|$). Graphs in panels c, f, and i show the results of a variation partition after fitting data to the HMSC statistical model. The variation partition plot shows the proportion of variation explained by environment, trait evolution ($|\\Delta x_1|$ and $|\\Delta x_2|$), and temporal and spatial random effects for $ln(N)$ for species 1 and 2. Error bars and shading in panels a, d, and g represent the 95% highest probability density interval for posterior predictions."), echo=FALSE}
knitr::include_graphics("../output/fig1_mod.pdf",error=FALSE)
```

@@

The simulation model followed population size and trait value dynamics for species in the landscape across 200 time steps (Figure \@ref(fig:fig2) a). Resulting biodiversity dynamics depended on the dispersal level and heritability values (Figure \@ref(fig:fig2) b). For dispersal, results mirrored some expectations from existing metacommunity models, i.e. that regional $\gamma$ diversity decreases with increasing dispersal level $d$ ([@Mouquet2003]; Figure \@ref(fig:fig2) b). The inclusion of trait heritability had a strong effect on community diversity, reflecting that local adaptation can rescue species and contribute to richness at the local and metacommunity scale. Regional species diversity ($\gamma$) increased with increasing $h^2$ levels (Figure \@ref(fig:fig2) b).

Analysis using the HMSC model for community time series across $d$ and $h^2$ levels indicated that the importance of evolution for community composition ($V_{|\Delta x|}$) decreased with increasing dispersal and with increasing heritability (Figure \@ref(fig:fig2) c). The importance of spatial random effects increased with dispersal rate, and overall the importance of evolution decreased with increasing $h^2$. These effects can be better understood by our supplemental analyses of 1-species, 1-site simulation models of evolutionary rescue (Appendix S3).

```{r fig2, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("Importance of trait evolution for change in community composition in metacommunities. (a) We developed a simulation model for population growth, competition, and trait evolution in 15 species that inhabit a patchy landscape. The landscape has 50 sites (colored points in the XY space), with distinct local environments that select for diverse patterns of species diversity (size of points) and average trait values (color of points; results are shown at time $t=0, 10, 200$ for a simulation with $d=10^-3$ and $h^2 = 0.1$). (b) Regional metacommunity $\\Gamma$ diversity varied across levels of dispersal and heritability ($h^2$). (c) Variation partition plots after fitting metacommunity time series of species abundances $ln(N_{ikt})$ to HMSC models with environment, trait evolution, and random effects (space and time) as predictors. The top panel is for $h^2 = 0.1$, the middle panel is for $h^2 = 0.3$, and the bottom panel is for $h^2 = 0.9$."), echo=FALSE}
knitr::include_graphics("../output/fig2_mod.pdf",error=FALSE)
```

@@

The results for the HMSC analysis of the empirical data (community dynamics and frond size of four aquatic plant species under different environmental conditions) estimated an effect of trait change for species density. We used the HMSC models to calculate the mean proportion of variance explained (variance partitioning, VP) over all species, to estimate the relative importance of evolutionary (change in the frond size trait) and ecological (light and nutrient treatments) effects on species densities.Almost one fifth (18.6%) of the variance in species density could be explained by the variance in $|\Delta x|$ (\@ref(fig:fig3)a). Including trait evolution in the statistical model accurately led to reductions for the estimated importance of ecological effects by a total of 14.4% (from 37% to 26.6% for the amount of shading and from 31% to 27.1% for the nutrient concentrations) and the reduction of 4.2% of the random effects (from 18.7% to 15% for time and from 13.3% to 12.8% for replicates) (\@ref(fig:fig3)). While $|\Delta x|$ of the species Lm, Sp and Wc showed a similarly low impact (3.7%, 3.4%, and 3.7% respectively), $|\Delta x|$ of species Rn showed a higher impact on species density, 7.8%, which can be explained by its high $|\Delta x|$ changes over time (see Appendix S4 data). The 95% HPDI for main effect sizes showed a negative effect of light shading for species density, while nutrient concentration had a positive estimated effect for density (\@ref(fig:fig3)c). Changes in $|\Delta x|$ impacted densities in distinct ways for intra- and interspecific comparisons. The frond size of species Rn had a negative impact on Rn density, while change in Wc frond size had a positive estimated effect on species Sp and Lm (\@ref(fig:fig3)c). We show the estimated effect of trait evolution for species counts in \@ref(fig:fig3)d, which reflects potentially strong impacts in particular species, density, and light treatments and weak effects with high uncertainty for many cases.

```{r fig3, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("The estimated effects of species trait changes and environmental treatments for densities of experimental aquatic plants (species Lm, Rn, Sp, and Wc). The variation partition for the HMSC model with (a) and without (b) trait change $|\\Delta x|$ is shown, as well as (c) the posterior estimated direction of effect (negative or positive; effect sizes with 95% HPDI that included 0 are colored white). The results indicated negative impacts of light, positive impacts of nutrients, negative intraspecific effects of $|\\Delta x_{Rn}|$ for $log(N_{Rn})$, negative interspecific effects of $|\\Delta x_{Lm}|$ for $log(N_{Sp})$, and positive interspecific effects of $|\\Delta x_{Wc}|$ for $log(N_{Sp})$ and $log(N_{Lm})$. (d) The posterior effect size estimates were used to estimate model-predicted densities at each time step of the experiment as a function only of $|\\Delta x|$ (for each of the four species, averaged across all of the other treatment effects). This captures the expected change in density due to trait evolution in all species."), echo=FALSE}
knitr::include_graphics("../output/fig3.pdf",error=FALSE)
```

@@

The analysis of community composition often requires statistical models that can consider the combined influences of environmental drivers and spatial connectivity. A more difficult, yet critical, element to incorporate is species interactions. Although the causal role of species interactions is difficult to interpret from statistical models (but see [@Dubart2022] [@Luo2022]) in systems with few species), joint species distribution models (JSDMs) have implemented spatial and temporal random effects that can effectively capture these species interactions using a latent variable approach (but note the resulting species covariance matrices will also reflect other processes besides species interactions; [@Poggiato2021]). Finally, one element that has not been fully implemented into joint species distribution models is dynamic trait evolution (see [@Garzon2019] for a model that does consider reaction norms from in-situ reciprocal transplant experiments). We used an existing JSDM, and developed simulation models of evolving populations in increasing complex scenarios to determine whether the signal of trait evolution could be estimated from data and compared to other drivers of community structure. We also applied this statistical model (HMSC; [@Ovaskainen2017HMSC] [@Tikhonov2020]) to an existing empirical dataset (from the experimental study of community eco-evolutionary dynamics in aquatic plants of [@Jewell2023]) to show how HMSC can be used to model the changes in species composition over time as a function of environmental drivers, trait change, and temporal random effects. Overall, there is no barrier to including dynamic trait change as a fixed effect in a JSDM model, and the resulting effect size coefficient estimates can be used to (i) estimate the relative importance of trait evolution for community composition and (ii) predict the impacts of future trait change in the system.

>>>>

Biodiversity at the metacommunity scale is typically influenced by a number of environmental, spatial, biotic, and stochastic factors. At the same time, these factors impact the evolution of individual species, as sites present different local selection pressures and connectivity can impact gene flow and genetic drift. Identifying the relative impacts of environmental, spatial, biotic, and other drivers on community composition across spatial and temporal scales has been greatly facilitated by joint species distribution models, but these models have yet to consider the impact of microevolution on community composition. We used Heirarchical Models of Species Communities (HMSC) to analyze simulated data of an populations and communities, including in a large evolving metacommunity model, to establish whether HMSC can sufficiently quantify the contribution of phenotypic evolution for metacommunity composition. The models successfully partitioned variance contributed by environmental, spatial, and evolving phenotypic drivers, and also estimated site- and year-specific covariance. We also applied the HMSC with trait evolution model to an existing dataset studying trait change and community dynamics in an experimental aquatic plant system. The study of eco-evolutionary dynamics may require data that reflects numerous complex, interacting processes and it is necessary to have flexible, generalized statistical models to analyze this data. JSDM models such as HMSC present one promising path for analysis of eco-evolutionary dynamics in multi-species communities.

@@

We evaluated whether linear regression models can appropriately capture the dynamics that drive species abundances in relatively simple populations and communities. We fit data generated from simulation models with increasing complexity (Table \@ref(tab:tab1)) to Bayesian linear regression models. These simulation models were (A) a single-species model of discrete-time logistic population growth, (B) expanded to include a randomly fluctuating environmental property $E$ that impacts the species' population growth rate, (C) a two-species model of logistic growth and competition that was then (D) expanded to consider a randomly fluctuating environmental property, (E) and further expanded to consider evolution in a trait $x$ that is selected by the environmental property, and finally (F) the model of two-species growth, competition, environmental change, and evolution in a spatially structured (10-patch) habitat.

@@

To fit linear regression models to the simulated population size data, we began with models using population size (model *A*) and environment (including the second-order term, to better reflect species optimum values; model *B*) as fixed effects ($N_t \sim N_{t+1} + E_t + E_t^2$), and progressed to a mixed model using environment and trait evolution as fixed effects ($N_t \sim E_t + E_t^2 + |\Delta x|$; for simulation model *E*) and time (models *C*, *D*, *E*, and *F*) and spatial distance (model *F*) as random effects. For consistency, all linear regression models were implemented as hierarchical Bayesian models using HMSC (Hierarchical Modelling of Species Communities [@Ovaskainen2017HMSC] [@Tikhonov2020]). HMSC uses a latent variable approach as an alternative to directly modelling species interaction coefficients, estimating a reduced number of linear combinations of species abundances that best predict future population size values. For linear models fit to data generated in simulations *C*, *D*, *E*, and *F* above, the time random effect was implemented to capture species interactions (as part of the temporal species association matrix estimated by the model), and the spatial random effect was implemented in model *F* to capture spatial covariance in abundances as a function of distance between sites. Full methodological details are given in Appendix S1. The one-two species simulations were run in R [@R2024] using the `ecoevoR` package [@Pantel2023] and HMSC models were implemented using the `Hmsc` package [@Hmsc2022].

@@

All simulations used the same parameter values ($W_{max}$ = 2, *P* = 1, *w* = 2, $N_c$ = 100, *p* = 0.001), species interaction matrix ($\alpha_{ii} = 0.00125$; $\alpha_{ij} \sim U(0,0.0015)$), and initial conditions (initial richness $s_{0k} \sim Pois(0.75)$; initial population size $N_{0i} \sim Pois(10)$; initial degree of maladaptation $\beta_{i0} \sim Gamma(0.75,1)$, initial distance to the local patch's optimum phenotype $d_{i0}=\sqrt{B_{i0}(w+P)}$, initial trait value $x_{i0} = E_k - d_{i0}$; Appendix S4). Simulations were run across a range of dispersal values (identical for all species; *d* = 0, $10^{-9},10^{-8},10^{-7},10^{-6},10^{-5},10^{-4},10^{-3},10^{-2}$, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) and heritability values (identical for all species: $h^2$ = 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) for 200 time steps, producing records of $N_{ikt}$ and $x_{ikt}$. Code for all simulations and model fits is available at [https://github.com/jhpantel/ecoevo-hmsc/tree/aRxiv](https://github.com/jhpantel/ecoevo-hmsc/tree/aRxiv).

@@

To determine the relative influence of environmental and spatial properties, intra- and interspecific population dynamics, and evolutionary dynamics for community structure, we applied HMSC to model the log-abundance at each site and year for each species. The log-abundance values were modeled as $ln(N_{ikt}) \sim N(L_{ikt},\sigma_2)$, where the linear predictors **L** are the sum of fixed and random effects **L^F^** + **L^R^** [@Tikhonov2020]. The fixed effects were $E_k$, $E_k^2$, and the absolute value of the change in each species' trait value from the previous time step to the next $|\Delta x_{ik,t-1 \to t}|$. Random effects included time (the day of the sample) and space. The spatial latent variables represent a spatial model, where species respond to some latent spatial predictor associated with the *xy* coordinates of the sites [@Tikhonov2020]. When there was no dispersal in the system ($d=0$), spatial random effects were not included in the model.

### Applying HMSC to empirical data

To test the applicability for HMSC on ecological and evolutionary effects in empirical data, we applied HMSC to the experimental data of Jewel & Bell (2023) [@Jewell2023]. The authors conducted a community selection experiment in which they grew four floating and short-lived aquatic plants (the angiosperms *Lemna minor* (Lm), *Spirodela polyrhiza* (Sp) and *Wolffia columbiana* (Wc), and the liverwort *Ricciocarpos natans* (Rn)) under different environmental treatments (shade cover and nutrient levels) for twelve weeks. The shade cover and nutrient levels for the experimental pools were a factorial combination of 3%, 12% and 50% shade, and low (dissolved nitrogen (DN) = 200 $\mu g L^{-1}$, dissolved phosphorous (DP) = 10 $\mu g L^{-1}$), medium (DN = 800 $\mu g L^{-1}$, DP = 40 $\mu g L^{-1}$), and high (DN = 3200 $\mu g L^{-1}$, DP = 160 $\mu g L^{-1}$) concentrations of nutrients, respectively. The frond size of ten individuals from each species and the density of each species were measured weekly for the eleven weeks of the experiment. To estimate the importance of trait change in the HMSC model, we calculated the average trait change as $\Delta x = \bar x_{ijt} - \bar x_{ijt-1}$, where $\bar x_{ij}$ described the mean trait of species $i$ in environment $j$ and week $t$.

To determine the relative importance of trait change for community composition, we fit two HMSC models to the data: An 'evolution model' in which $|\Delta x|$ of each species and environmental treatment were used as explanatory variables, and a 'no-evolution model', in which only the environmental conditions were included. Species density was log-transformed species, and we included replicates (*n* = 2) and time as random effects. For each model we ran four MCMC chains to generate 1000 samples of the posterior distribution for all model parameters. To evaluate the model-estimated role of trait change for predicted population size, we used the posterior distributions of the estimated effect size of $|\Delta x|$ multiplied by the observed $\Delta \bar x$ of each species, added to the model-estimated intercept.

@@

Researchers often use linear regression models to study non-linear population and community dynamics. We evaluated the performance of linear models in simple cases (1-2 species, 1-few sites), to validate that effect size estimates and predicted densities accurately reflected mechanisms before moving onto more complex models in the next sections of our analysis. When only one species was present (models *A* and *B*), the regression coefficients (model *A*: $N_{t+1} \sim \beta_0 + \beta_1 N_t + \epsilon$; *B* $N_{t+1} \sim \beta_0 + \beta_1 N_t + \beta_2 E_t + \beta_3 E_t^2 + \epsilon$) reflected the direction and magnitude of the mechanistic drivers of population size. $\beta_1$ correctly captured the positive relationship between $N_t$ and $N_{t+1}$ under our simulation conditions, and $\beta_2$ and $\beta_3$ correctly captured the quadratic unimodal relationship between $E_t$ and $N_t+1$ (Appendix S1.1, S1.2).

When a second species was added (models *C* and *D*), the regression coefficients performed well for predicting density values at all time points, but the direction and magnitude of the regression coefficients did not reflect the intra- and interspecific competition strengths included in the mechanistic model (Appendix S1.3). We instead used a latent variable modelling approach (implemented in HMSC) to consider these species interactions as part of the temporal random effects included in the model. This approached produced effect size estimates that both predicted the population dynamics well (Figure \@ref(fig:fig1)a) and captured the conditional effects of environment for species densities (Figure \@ref(fig:fig1)b). The variation partition also successfully captured the relative importance of the environmental predictor and the temporal random effects for model *D* (which in this simple case were driven by intra- and interspecific species interactions; Figure \@ref(fig:fig1)c; Appendix S1.3, S1.4).

Trait evolution was included as a predictor when it was introduced in model *E*. The resulting linear regression coefficients could predict species dynamics (Figure \@ref(fig:fig1)d) and reflected the impact and magnitude of environmental and trait evolution predictors (Figure \@ref(fig:fig1)e). The variation partition also showed that trait evolution in species 1 ($|\Delta x_1|$) was a strong driver of variation in abundance (blue portion of bar plot, Figure \@ref(fig:fig1)f), primarily for species 1 but also for species 2 as well (Appendix S1.5; Appendix S2).

After adding spatial complexity to the 2-species simulation, we included a spatial random effect to capture species covariances as a function of distance between sites (model *F*). By properly considering species interactions as a temporal random effect and including spatial random effects, the HMSC model could estimate regression coefficients for the impacts of the environment $E_t$ and trait evolution $|\Delta x_1|$ and $|\Delta x_2|$ for both species that fit well to the simulated data points (Figure \@ref(fig:fig1)g); Appendix S1.6) and capture conditional effects of environment and trait evolution for both species (Figure \@ref(fig:fig1)h). Spatial random effects explained a large proportion of variation in abundance for both species, but trait evolution remained a strong driver for species 1.

```{r fig1, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("Simulation models for 2-species ecological and eco-evolutionary dynamics. Panels a, d, and g show log-population size (y-axis) over time (x-axis) for species $1$ and $2$ when they grow and compete for resources with random environmental fluctuations over time. Error bars give the 95% highest probability density interval (HPDI) for posterior predictions from the HMSC model fit. Species either have fixed (panel a) or evolving trait values (panels d, g) and are present in either 1 (panels a, d; simulation models D and E) or multiple sites (panel g; model F). Graphs in panels b, e, and h show observed (points)  and model-predicted (black line with blue shading) log-population size ($log(N)$) for species 1 and 2 across gradients of the environment ($E + E^2$) and the change in trait values of species 1 ($|\\Delta x_1|$). Graphs in panels c, f, and i show the results of a variation partition after fitting data to the HMSC statistical model. The variation partition plot shows the proportion of variation explained by environment, trait evolution ($|\\Delta x_1|$ and $|\\Delta x_2|$), and temporal and spatial random effects for $ln(N)$ for species 1 and 2. Error bars and shading in panels a, d, and g represent the 95% highest probability density interval for posterior predictions."), echo=FALSE}
knitr::include_graphics("../output/fig1_mod.pdf",error=FALSE)
```

@@

The simulation model followed population size and trait value dynamics for species in the landscape across 200 time steps (Figure \@ref(fig:fig2)a). Resulting biodiversity dynamics depended on the dispersal level and heritability values (Figure \@ref(fig:fig2)b). For dispersal, results mirrored some expectations from existing metacommunity models, i.e. that regional $\gamma$ diversity decreases with increasing dispersal level $d$ ([@Mouquet2003]; Figure \@ref(fig:fig2)b). The inclusion of trait heritability had a strong effect on community diversity, reflecting that local adaptation can rescue species and contribute to richness at the local and metacommunity scale. Regional species diversity ($\gamma$) increased with increasing $h^2$ levels (Figure \@ref(fig:fig2)b).

Analysis using the HMSC model for community time series across $d$ and $h^2$ levels indicated that the importance of evolution for community composition ($V_{|\Delta x|}$) decreased with increasing dispersal and with increasing heritability (Figure \@ref(fig:fig2)c). The importance of spatial random effects increased with dispersal rate, and overall the importance of evolution decreased with increasing $h^2$. These effects can be better understood by our supplemental analyses of 1-species, 1-site simulation models of evolutionary rescue (Appendix S3).

```{r fig2, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("Importance of trait evolution for change in community composition at the metacommunity scale. (a) We developed a simulation model for population growth, competition, and trait evolution in 15 species that inhabit a patchy landscape. The landscape has 50 sites (colored points in the XY space), with distinct local environments that select for different patterns of local species diversity (size of points) and average trait values (color of points; results are shown at time $t=0, 10, 200$ for a simulation with $d=10^{-3}$ and $h^2 = 0.1$). (b) Regional metacommunity $\\gamma$ diversity decreased with increasing disperal level and increased with increasing heritability ($h^2$). (c) Variation partition plots after fitting metacommunity time series of species abundances $log(N_{ikt})$ to HMSC models with environment, trait evolution, and random effects (space and time) as predictors. These variation partition results are averaged across all species in the metacommunity. The top panel is for $h^2 = 0.1$, the middle panel is for $h^2 = 0.3$, and the bottom panel is for $h^2 = 0.9$. The alluvial plots show the change in relative importance of different predictors for increasing dispersal levels."), echo=FALSE}
knitr::include_graphics("../output/fig2_mod.pdf",error=FALSE)
```

@@

We fit the empirical data (community dynamics and frond size of four aquatic plant species under different environmental treatments) to an HMSC model, and estimated the effect size of shade cover, nutrient level, and trait change (in all four species) for changes in species densities. The variation partition produced by the HMSC model fit shows the estimated relative importance of evolutionary (change in the frond size trait) and ecological (light and nutrient treatments) effects on species densities. Almost one fifth (18.6%) of the variation in species density could be explained by the variance in $|\Delta x|$ (Figure \@ref(fig:fig3)a; dLm, dRn, dSp, dWc). Including trait change in the statistical model accurately led to reductions for the estimated importance of ecological effects by a total of 14.4% (from 37% to 26.6% for the amount of shading and from 31% to 27.1% for the nutrient concentrations) and the reduction of 4.2% of the random effects (from 18.7% to 15% for time and from 13.3% to 12.8% for replicates) (Figure \@ref(fig:fig3)). While $|\Delta x|$ of the species Lm, Sp and Wc showed a similarly low impact (3.7%, 3.4%, and 3.7% respectively), $|\Delta x|$ of species Rn showed a higher impact on species density, 7.8%, which can be explained by its high $|\Delta x|$ changes over time (see Appendix S4 for data). The 95% HPDI for main effect sizes showed a negative effect of light shading for species density, while nutrient concentration had a positive estimated effect (Figure \@ref(fig:fig3)c). Changes in $|\Delta x|$ impacted densities in distinct ways for intra- and interspecific comparisons (Figure \@ref(fig:fig3)c). The change in frond size of species Rn (dRn) had a negative impact on Rn density, while change in Wc frond size (dWc) had a positive estimated effect on species Sp and Lm (Figure \@ref(fig:fig3)c). We show the estimated effect of trait change for species counts in Figure \@ref(fig:fig3)d, which reflects potentially strong impacts in some particular species and environmental treatment combinations, and weak effects with high uncertainty for many species and environmental treatments.

```{r fig3, out.width="110%", include=TRUE, fig.align="center", fig.cap=c("The estimated effects of species trait changes and environmental treatments for densities of experimental aquatic plants (species Lm, Rn, Sp, and Wc). The variation partition for the HMSC model with (a) and without (b) trait change $|\\Delta x|$ is shown, as well as (c) the posterior estimated direction of effect (negative or positive; effect sizes with 95% HPDI that included 0 are colored white). The model estimated negative impacts of light, positive impacts of nutrients, negative intraspecific effects of $|\\Delta x_{Rn}|$ for $log(N_{Rn})$, negative interspecific effects of $|\\Delta x_{Lm}|$ for $log(N_{Sp})$, and positive interspecific effects of $|\\Delta x_{Wc}|$ for $log(N_{Sp})$ and $log(N_{Lm})$. (d) The posterior effect size estimates were used to predict species densities at each time step of the experiment. Here we show results only for the main effects of $|\\Delta x|$ (for each of the four species, averaged across all of the other treatment effects). This captures the expected change in density due to trait evolution in all species."), echo=FALSE}
knitr::include_graphics("../output/fig3.pdf",error=FALSE)
```

@@

The analysis of community composition often requires statistical models that can consider the combined influences of environmental drivers and spatial connectivity. A more difficult, yet critical, element to incorporate is species interactions. Although the causal role of species interactions is difficult to interpret from statistical models (but see [@Dubart2022] [@Luo2022]) in systems with few species), joint species distribution models (JSDMs) have implemented spatial and temporal random effects that can effectively capture these species interactions using a latent variable approach (however it is important to note that the resulting species covariance matrices also reflect other processes besides species interactions, including unmeasured environmental effects; [@Poggiato2021]). One element that has not been fully implemented into joint species distribution models is dynamic trait evolution (see [@Garzon2019] for a model that does consider reaction norms from in-situ reciprocal transplant experiments). We used an existing JSDM, and developed simulation models of evolving populations in increasingly complex scenarios to determine whether the signal of trait evolution could be estimated from data and compared to other drivers of community structure. We also applied this statistical model (HMSC; [@Ovaskainen2017HMSC] [@Tikhonov2020]) to an existing empirical dataset (from the experimental study of community eco-evolutionary dynamics in aquatic plants of [@Jewell2023]) to show how HMSC can be used to model and predict the changes in species composition over time as a function of environmental drivers, trait change, and temporal random effects. Overall, there is no barrier to including dynamic trait change as a fixed effect in a JSDM model, and the resulting effect size coefficient estimates can be used to (i) estimate the relative importance of trait evolution for community composition and (ii) predict the impacts of future trait change in the system.
